---
title: "Crimes and incarceration in the United States"
author: "Ewelina Osowska and Ewa Sobolewska"
date: "10-01-2019"
output:
  rmarkdown::html_document:
    theme: yeti
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(gridExtra)
library(ggplot2)
library(plotly)
library(panelr)
library(knitr)
library(funModeling)
library(spData)
library(sf)
library(tmap)
library(kableExtra)
library(formattable)

kable_f = function(...) {
  knitr::kable(..., format.args = list(decimal.mark = '.', big.mark = ",")) %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F)
}
```

# Introduction

blablabla

# Data description and preprocessing

The datasets used in the below analysis were sourced from www.kaggle.com website [^kaggle]. They were created based on several sources including the Bureau of Justice Statistics [^bjs] and FBI Uniform Crime Reporting Program [^fbi]. The National Prisoner Statistics Program conducted by the Bureau of Justice Statistics has collected data on the number of prisoners in state and federal prison facilities since 1926. It is produced annually on national and state level. Data are sourced from the 50 state departments of correction, the Federal Bureau of Prisons, and until 2001, from the District of Columbia. The UCR Program provides statistics on violent and property crimes. Data are collected annually and are available on national, state and city level. For the purposes of our analysis we are using state-level statistics.

Additionally, we individually collected data on prison expenditures provided by the Bureau of Justice Statistics [^prison_exp] for each state in 2016 which is the lastest data available. Later in the analysis we will use them in order to correlate the expendirutes with the occurence of particular crimes.

## UCR 

The UCR dataset consist of 15 variables, two of which are the jurisdiction and year of the observation. It provides information about the state population and also about number of violent crimes (murder, manslaughter, rape, robbery, aggravated assault) and property crimes (burglary, larceny, vehicle theft) per state yearly. Detailed definitions of each crimes can be found on UCR Program website.

The *crime_reporting_change* variable reflects instances when states' reporting standards changed. The *crimes_estimated* variable indicates cases where the FBI computes estimates for participating agencies not providing 12 months of complete data for state [^estimated_crime].

```{r, message=FALSE, warning=FALSE}
ucr <- read_csv("data/ucr_by_state.csv")
ucr$year <- as.factor(ucr$year)
```

```{r, message=FALSE, warning=FALSE}
colnames(ucr)
```

The ucr dataset has a lot of missing values, compared to the other datasets that have none. We dropped the last 6 columns that were completely empty and then we dropped rows consisting of only missing values. It leaves all columns without any missing values apart from "rape_revised" with 612 missing values and "rape_legacy" with 104 missing values.

```{r}
# removing last 6 columns
ucr <- ucr[, -c(16:21)]
# removing all missing rows
ind <- apply(ucr, 1, function(x) all(is.na(x)))
ucr <- ucr[ !ind, ]
# showing sum of missing values per columns
sapply(ucr, function(x) sum(is.na(x)))
```

As you can see on plot on the left below, in the last two years, 2016 and 2017, there is an additional obervation ie. jurisdiction. Looking at the plot on the right, New York is missing in one year, Puerto Rico is visible in only 3 years. District of Columbia is sometimes renamed as DC, but overall it sums up to all 17 years.

```{r, fig.width = 10, fig.asp = .25, message=FALSE, warning=FALSE}
library(viridis)
plot.data1 = ucr %>% group_by(year) %>% count()
ggp1 = ggplot(data = plot.data1, aes(x=year, y=n, fill=year)) + 
  geom_bar(stat = "identity") +
  scale_fill_viridis_d() +
  scale_x_discrete(breaks = as.factor(seq(2001, 2017,2))) +
  theme_minimal() + 
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank(),
        legend.position = "none")

plot.data2 = ucr %>% group_by(jurisdiction) %>% count() %>% arrange(n) %>% filter(n<17)
ggp2 = ggplot(data = plot.data2, aes(x=jurisdiction, y=n, fill=jurisdiction)) + 
  geom_bar(stat = "identity") +
  theme_minimal() + 
  scale_fill_viridis_d() +
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank(),
        legend.position = "none")

grid.arrange(ggp1, ggp2, ncol = 2)
```

Based on the above analysis, we decided to rename "DC" to "District of Columbia" and exclude Puerto Rico state.

```{r}
ucr$jurisdiction[ucr$jurisdiction=="DC"] <- "District of Columbia"
ucr <- ucr %>% filter(jurisdiction!="Puerto Rico")
```

```{r}
# interpolation <- data %>%
#   group_by(country) %>%
#   mutate(valueIpol = approx(year, women_part, year, 
#                             method = "linear", rule = 1:2, f = 0, ties = mean)$y)
# i=0
# for (i in seq_along(interpolation$valueIpol)) {
#   if (is.na(interpolation$women_part[i]) == FALSE) 
#     i = i+1
#   else if (is.na(interpolation$women_part[i]) == TRUE) 
#     interpolation$women_part[i] <- interpolation$valueIpol[i]
# }
```

We also analysed the missing values of variables *rape_revised* and *rape_legacy*. Because there are so many missings and they mostly do not occur in the same year, we can't compare them and that's why we decided to drop them. 

```{r}
rape_df <- data.frame(year=as.factor(2001:2017))

rape_revised_count <- ucr[!is.na(ucr$rape_revised),] %>% 
                            group_by(year) %>% 
                            count(name="rape_revised_count")
rape_legacy_count <- ucr[!is.na(ucr$rape_legacy),] %>% 
                            group_by(year) %>% 
                            count(name="rape_legacy_count")
rape_df <- left_join(rape_df, rape_revised_count, by="year")
rape_df <- left_join(rape_df, rape_legacy_count, by="year")
```
### {.tabset .tabset-fade .tabset-pills}
#### Hide data

#### Show data
```{r}
kable_f(rape_df)
```
### <span style="color: white; font-size:8px;">.</span>
```{r, message=FALSE, warning=FALSE}
ucr$rape_legacy <- NULL
ucr$rape_revised <- NULL
```

```{r}
colnames(ucr)
```

```{r, fig.width = 7, fig.asp = 1.1,}
pl <- vector("list", length = ncol(ucr[,c(5:13)])-1)
colors <- viridis(8)
for(ii in seq_along(pl)){
  .col <- colnames(ucr[,c(5:13)])[-1][ii]
  .p <- ggplot(ucr, aes_string(x=.col, fill="colors[ii]", color="colors[ii]")) + 
          geom_density(alpha=0.3) + 
          scale_fill_manual(values = colors[ii], aesthetics = c("color", "fill")) +
          theme_minimal() +           
          theme(legend.position = "none",
                axis.title.x = element_blank(),
                axis.title.y = element_blank()) +
          labs(title = .col)+ 
          scale_y_continuous(labels = function(x) format(x, scientific = FALSE))+ 
          scale_x_continuous(labels = function(x) format(x, scientific = FALSE))
  
  pl[[ii]] <- .p
}

grid.arrange(grobs=pl, ncol=2)
```
 (...)

## Incarcarations in prison

```{r, message=FALSE, warning=FALSE}
prison <- read_csv("data/prison_custody_by_state.csv")
colnames(prison)
```

The prison data, compared to ucr is in a panel form, consisting of years as columns. Using long_panel we converted the dataframe so that each row is a different jurisdiction and year.

```{r}
colnames(prison)[3:18] <- paste0(colnames(prison)[3:18],'1')
prison_panel <- long_panel(prison, begin = 2001, end = 2016, label_location = "beginning", id = "jurisdiction")
names(prison_panel)[names(prison_panel) == "wave"] <- "year"
names(prison_panel)[names(prison_panel) == "1"] <- "prison"
prison_panel$year <- as.factor(prison_panel$year)
kable_f(head(prison_panel))
```


## Prison expenditures

```{r, message=FALSE, warning=FALSE}
prison_exp_2016 <- read_delim("data/prison_expenditures.csv", ";")
```

```{r, message=FALSE, warning=FALSE}
kable_f(head(prison_exp_2016))
```

## State area and region

In order to enhance further visualisations, we add an information about state area and region based on R built-in *us_states* dataset. 

```{r warning=FALSE}
library(spData)
library(sf)
us_states_info <- data.frame(jurisdiction = us_states$NAME, 
                             region = us_states$REGION,
                             area_km2 = as.numeric(round(us_states$AREA, 0)))

kable_f(head(us_states_info))
```

Because of the fact that there are two states missing in the us_states_info dataset, we manually added region and land area for Hawaii and Alaska [^area].

```{r warning=FALSE}
additional_states <- data.frame(jurisdiction = c("Hawaii", "Alaska"),
           region = c("remote", "remote"),
           area_km2 = c(16638, 1481346))

us_states_info <- rbind(us_states_info, additional_states)
```

## Unifying state names

In the prison dataset, District of Columbia is named as Federal and in prison_exp_2016 is named as Washington, D.C., so in order to unify the names we ranamed both to District of Columbia. We also renamed the variable State and type of government to jurisdiction for easier further calculations.

```{r}
setdiff(prison$jurisdiction %>% unique(), ucr$jurisdiction %>% unique())
```

```{r}
setdiff(prison_exp_2016$`State and type of government` %>% unique(), ucr$jurisdiction %>% unique())
```

```{r}
setdiff(ucr$jurisdiction %>% unique(), us_states_info$jurisdiction %>% unique())
```

```{r}
prison$jurisdiction[prison$jurisdiction=="Federal"] <- "District of Columbia"
names(prison_exp_2016)[names(prison_exp_2016) == "State and type of government"] <- "jurisdiction"
prison_exp_2016$jurisdiction[prison_exp_2016$jurisdiction=="Washington, D.C."] <- "District of Columbia"
```

# Background

The United States has the largest prison population in the world, and the highest per capita incarceration rate. According to 2018 report of the Bureau of Justice Statistics (BJS), nearly 2.2 million adults were imprisoned in America at the end of 2016. That means for every 100,000 people living in the US, about 655 of them were held in prisons and jails. Because of the huge scale of prisoners in the country, also the expenditures on prisons are the highest. According to recent surveys regarding the United States expenditures, spendings on incarceration have increased about three times as fast as spendings on elementary and secondary education during this time period. 

Accorindg to Hartney [^hartney], 



The incarceration rate in the US is four times the world average.
Some individual US states imprison up to six times as many people as do
nations of comparable population.
The US imprisons the most women in the world.
Crime rates do not account for incarceration rates.

(...)


https://www.nccdglobal.org/sites/default/files/publication_pdf/factsheet-us-incarceration.pdf



1. liczba prisoners:

```{r, warning=FALSE}
prison_panel_year <- prison_panel %>% group_by(year) %>% summarise(value = sum(prison))

p <- ggplot(data = prison_panel_year, aes(x = year, y = value, color = year)) +
  geom_point() +
  scale_color_viridis_d() +
  labs(title = "Number of prisoners in the USA per year", x = "Year", y = "Number of prisoners") +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p, tooltip = "text")
```

### Average historical violent crimes vs. property crimes per population across states

Below can be seen maps of US states divided by the severity of a crime, that is violent compared to property crimes.

```{r, fig.width = 10}

ucr %>% left_join(us_states_info, by="jurisdiction") %>% 
          group_by(region) %>% 
          summarise(violent_crime_total = mean(violent_crime_total),
                    property_crime_total = mean(property_crime_total)) %>%  kable_f()

# ggplot(data=left_join(ucr, us_states_info, by="jurisdiction")) + geom_density(aes(x=property_crime_total))+ geom_density(aes(x=violent_crime_total))+ facet_grid(rows = vars(region))
```


```{r}
#create df with mean values across years per state from ucr
ucr_grouped <- ucr %>% 
                  group_by(jurisdiction) %>% 
                  summarise(violent_crime_total = mean(violent_crime_total),
                            property_crime_total = mean(property_crime_total))
#rename variable for merging
names(ucr_grouped)[names(ucr_grouped) == "jurisdiction"] <- "NAME"
#merge grouped ucr and state spatial data
us_states_ucr <- merge(us_states, ucr_grouped, by = "NAME")

#create values per population
us_states_ucr$violent_crime_per_pop <- us_states_ucr$violent_crime_total/us_states_ucr$total_pop_15
us_states_ucr$property_crime_per_pop <- us_states_ucr$property_crime_total/us_states_ucr$total_pop_15

# create usa map for both crime types
usa1 <- ggplot(data = us_states_ucr) +
          geom_sf(aes(fill = property_crime_per_pop), lwd = 0) +
          scale_fill_viridis_c(option = "viridis", trans = "sqrt") +
          theme_minimal()
usa2 <- ggplot(data = us_states_ucr) +
          geom_sf(aes(fill = violent_crime_per_pop), lwd = 0) +
          scale_fill_viridis_c(option = "viridis", trans = "sqrt") +
          theme_minimal()

# format main map
usa_all1 <- usa1 + 
              ggtitle("Property crimes per population")+
              theme(legend.position = "right")
usa_all2 <- usa2 + 
              ggtitle("Violent crimes per population")+
              theme(legend.position = "right")
# zoom and format zoomed map of DC
usa_dc1 <- usa1 + 
            coord_sf(xlim = c(-79, -75), ylim = c(38, 40)) +
            guides(fill=FALSE)+
            theme(axis.title = element_blank(), 
                  axis.text  = element_blank(),
                  axis.ticks = element_blank())
usa_dc2 <- usa2 + 
            coord_sf(xlim = c(-79, -75), ylim = c(38, 40)) +
            guides(fill=FALSE)+
            theme(axis.title = element_blank(), 
                  axis.text  = element_blank(),
                  axis.ticks = element_blank())

# combine both plots and add red rectangle around zoomed area
ggp1 <- usa_all1 + 
          annotation_custom(ggplotGrob(usa_dc2), xmin= -80, ymax= 35)+
          geom_rect(aes(xmin = -79, xmax= -75, ymin=38, ymax = 40), size=1, fill=NA, color="red")

ggp2 <- usa_all2 + 
          annotation_custom(ggplotGrob(usa_dc2), xmin= -80, ymax= 35)+
          geom_rect(aes(xmin = -79, xmax= -75, ymin=38, ymax = 40), size=1, fill=NA, color="red")

```


### {.tabset .tabset-fade .tabset-pills}
#### Property crimes per population

```{r, echo=FALSE}
ggp1
```

#### Violent crimes per population

```{r, echo=FALSE}
ggp2
```

### jakis kolejny rozdział

3. 




# Statistical analysis of the dataset

jaka jest zależność między liczbą więźniów (prison) a wystąpieniami poszczególnych crime na przestrzeni lat (ucr)? 
czy wzrost uwięzionych zminiejsza odsetek jakiegoś typu przestępstw? czy może jest stały wzrost/spadek przestępstw? (geom line i geom smooth)

Does this significant investment into imprisonment improve public safety? wydatki na więzienia a wystąpienia przestępstw - ogółem i w kategoriach, w roku 2016 (najnowsze dane); source: https://www.bjs.gov/index.cfm?ty=dcdetail&iid=286


jak wygląda liczba uwięzionych na przestrzeni lat? dla całego kraju i dla poszczególnych stanów?




dodatkowe zmienne
-> area (ok) - w kodzie
-> wydatki na prisons (ok) - w excelu , 2016



-> co poza mapą i bombelkami?
- heatmapa
- 


-> https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/  

[^kaggle]: Source: https://www.kaggle.com/christophercorrea/prisoners-and-crime-in-united-states#ucr_by_state.csv
[^bjs]: Source: https://www.bjs.gov/index.cfm?ty=dcdetail&iid=269  
[^fbi]: Source: https://www.ucrdatatool.gov/Search/Crime/State/RunCrimeStatebyState.cfm  
[^prison_exp]: Source: https://www.bjs.gov/index.cfm?ty=dcdetail&iid=286
[^estimated_crime]: "For agencies supplying 3 to 11 months of data, the national UCR Program estimates for the missing data by following a standard estimation procedure using the data provided by the agency. If an agency has supplied less than 3 months of data, the FBI computes estimates by using the known crime figures of similar areas within a state and assigning the same proportion of crime volumes to nonreporting agencies." (cited from https://www.ucrdatatool.gov/faq.cfm)
[^area]: Sources: https://en.wikipedia.org/wiki/Alaska and https://en.wikipedia.org/wiki/Hawaii

